Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?

### 2024-04-10 tex/review/response_letter.tex

\textbf{General comment:} \textit{[...] The research results reported are too premature for publication. More work is needed to substantiate the conclusions in your manuscript.}

\auth Thank you for recognizing the merits of our submission and for the opportunity to revise it. In the following document, we provide detailed responses to the comments from the Reviewers and describe the changes made to the manuscript to address them. All corrections and clarifications to the manuscript are presented in \nt{red} color. The major changes are summarized below:

\begin{itemize}
  \item Reviewers 1 and 3 raised questions and requested clarifications about the number of scenarios used in the mFRR lookback strategy. We have made changes to the manuscript accordingly to explain our choices while also including an additional section in the appendix on the sensitivity analysis on the mFRR 2021 model (using ADMM as a solution strategy) of the number of scenarios.

  \item Reviewers 1 and 2  requested clarifications about the various energy prices used in the case studies, especially confusion regarding day-ahead (spot) prices and retail prices. This has lead to revisions to the manuscript to explicitly note the distinction and the Europe and the Nordics do not use time-of-use prices and consumers are directly exposed to spot prices. Furthermore, we changed figures 4 and 5 by merging them into one figure clearly explaining both the mFRR market and load shifting. Special focus has been devoted to making it clear that baseline power is indeed changed for load shifting but remains a parameter for the mFRR model.

@@

  \item Reviewer 2 requested clarifications regarding the phrase \textit{flexible load} we use to describe a load that can shift its power consumption in time. We agreed that it is too general and we have replaced the phrase with \textit{thermostatically controlled load} (or TCL) everywhere.
\end{itemize}

%%%%%%%%%%%%%%%%%
% REVIEWER 1
%%%%%%%%%%%%%%%%
\newpage
\section{Response to Reviewer 1}

\subsection{Summary Comment} \textit{This paper builds a mixed-integer linear program to maximize the flexibility value of the freezer and analyzes the demand-side flexibility provision in two forms of manual Frequency Restoration Reserve (mFRR) services and load shifting. However, there are some aspects of the content that require further explanation. The detailed comments are listed as follows.}

\auth We thank the Reviewer for thoughtful and constructive feedback on our manuscript. We appreciate the insights and carefully addressed the points raised to enhance the clarity and depth of our content. We have devoted our attention to justifying the number of scenarios used in our mFRR lookback strategy.

@@

Further, we address the Specific Comments in the following.

\subsection{Specific Comments}
\textbf{Comment 1:} \textit{Why did the assumption that the Balance Responsible Party is not uncertain about the mFRR market price? This assumption makes the model of load shifting deterministic, increasing the gap between the model and reality.}

\auth For load shifting, day-ahead prices are known beforehand and consumers are directly exposed to those prices. That is the case for Europe as opposed to North America where retailers might impose additional time-of-use prices. Thus, in Denmark, the electricity price is known 12-36 hours in advance which is in accordance with how we model load shifting. We have made changes to make it clear that day-ahead prices are known beforehand (please see below). Furthermore, the mFRR reservation market price is not uncertain in our model, but in reality it is (as a result of a uniform clearing auction). However, it remains very stable and the main source of uncertainty relates to the activation of mFRR which we indeed model. Hence, our goal is to compare mFRR with uncertain activation and load shifting with no uncertainty.

@@

\changes We have added the following sentence in the revised manuscript (p.9, Section 2.3) to clarify electricity prices used for load shifting:

\nt{For this, the TCL makes a load shifting decision right after the day-ahead market as shown by the blue box in Fig. 4. Note that in Denmark and generally in Europe, day-ahead prices are directly exposed to consumers with no fixed time-of-use price in between as is often the practice in North America.}

@@

\textbf{Comment 2:} \textit{There are different numbers of scenarios in the two mFRR cases, which would affect the comparability of the two cases. Please make some revisions to ensure that the comparison cases have the same number of scenarios.}

\auth Thanks for this astute comment. It is indeed a valid observation. First of all, the purpose of the ADMM strategy is to present a simple, i.e., fixed policy, that an aggregator can use to bid for entire 2022 without ever having to run a daily optimization. Such a policy is practical and appealing to real-life aggregators. On the contrary, the lookback policy might exhibit a better performance, especially in case of non-stationarity, but more complicated to implement and incorporate in daily operations. None of this was mentioned in our initial manuscript, hence we have added some sentences to clarify (please see below).

Second, during the analysis phase preceding this manuscript, a sensitivity analysis was made of both the ADMM strategy (based on scenarios from 2021) and for the lookback strategy (based on the last five days). The sensitivity analyses focused on the number of scenarios used in both cases and showed that ADMM performed the best when using at least 50 scenarios (with no improved performance using additional scenarios which we added as an appendix to the manuscript), and the same was the case for using five days in the lookback strategy. This has been clarified in the revised manuscript (please see below).

@@

\changes Regarding the first item mentioned with the real-life practicality of a fixed policy (ADMM solution strategy) vs the more complicated lookback strategy, we have added the following to (p.10, Section 3.2):

\nt{We use two different strategies for in-sample scenario generation: (\textit{i}) considering historical spot and balancing prices in DK2 in 2021.
%, with different cases where the number of scenarios $|\Omega|$ is 1, 5, 10, 20, 30, 40, 50, 100, and 250. 
(\textit{ii}) considering prices of the most recent five days (lookback strategy). This first scenario generation strategy represents the case where an aggregator relies on a simple, fixed policy without the need to run a daily optimization as opposed to the second  strategy, although it might exhibit a better performance in case the underlying uncertainty is non-stationary. The first  strategy uses 50 scenarios. Using a sensitivity analysis, we found out that increasing the number of scenarios beyond 50 does not improve the performance remarkably as explained in Appendix B. We made a similar observation for choosing five days in the second solution strategy.}

We then add Appendix B (including newly added Figure 8) as follows:

\nt{A sensitivity analysis was carried out to investigate the number of scenarios to use in the mFRR model using  2021 data. Fig. 8 shows the in-sample (IS) and out-of-sample (OOS) costs when increasing the number of scenarios used from 2021. It was eventually chosen to use 50 scenarios as this provided a satisfactory performance both IS and OOS --- a higher number of scenarios did not yield any particular improvement.}

@@

\textbf{Comment 3:} \textit{Please add text annotations to the legend to improve the legibility of Fig.7 and Fig. 8. Additionally, does the "measurements" mean the real measurements in Fig. 7 and Fig. 8? Why is there a big difference between the curve of measurements and the curve of simulated results? Is this phenomenon due to inaccurate modeling or unreasonable assumptions? Could you narrow the difference?}

\auth Thanks for the comment. We have made changes to Figures 7 and 8 on the initial submission, accordingly, by adding descriptive names in parenthesis to the legends --- for example: $P_{h}^{\text{Base}}$ \textit{(Baseline power)}. The measurements are real-life data as shown in Figure 3. The point of the measurements in Figure 3 is to show how our state-space model is able to simulate reality, i.e., the measurements, to a satisfactory degree. In Section 2.1.3 (\textit{Model validation}), we have explained how our state-space model fits the measurements, and how it is able to simulate freezer temperatures for a whole day. To achieve a  satisfactory performance, i.e., a narrower difference to the measurements, requires even more sophisticated modeling techniques and is perhaps difficult to achieve given the current data provided. For example, there are a lot of uncertain events like people opening the lids to the freezers causing temperature drops. The important element of the state-space models used here is that they capture the most fundamental physics and dynamics from which one can directly use within an optimization model. However, we have conceded that it might be confusing to also show the measurements in Figures 7 and 8. We have therefore removed them from those figures.

@@

\changes The title of the manuscript has been changed as follows:

\nt{Load Shifting Versus Manual Frequency Reserve:  Which One is More Appealing to Thermostatically Controlled Loads in Denmark?}

\textbf{Comment 2:} \textit{Abstract: The use of a supermarket freezer as a representative flexible load is ok. But, trying to generalize the results for all types of flexible loads, as suggested by the title, is not ok. As mentioned above, flexible loads can include many types of loads. None of the other loads were studied in this paper. I strongly suggest replacing flexible loads with thermostatically controlled loads.}

\auth Thank you for your comment. As before, we do agree and have replaced all instances of \textit{flexible load} with \textit{thermostatically controlled load}.

@@

\textbf{Comment 3:} \textit{Overall paper: It is well written.}

\auth Thank you.

\textbf{Comment 4:} \textit{Section 1: It is not clear to me the context of load shifting in section 1. Is load shifting made considering spot energy prices or retail prices (potentially mimicking spot prices)? Please make this clear in section 1.}

\auth Thanks for the feedback. In Europe and in the Nordics, the day-ahead price (spot price) is directly exposed to consumers, i.e., there is no fixed time-of-use price in between via retailers. This was indeed not clear in our initial manuscript.

@@

\textbf{Comment 5:} \textit{Section 1.3: I struggled to identify novel research contributions, but I found the discussion of the paper interesting. For instance, there are papers in the literature that studied the participation of aggregators of prosumers (with flexible loads) in both energy and tertiary reserve (mFRR) markets using two-stage stochastic optimization models. My point is that this type of study exists in literature, however in different contexts and markets. This deserves to be discussed in the paper.}

\auth It is indeed the case that similar studies exist albeit under different settings and with different assumptions, especially around the level of knowledge assumed from an aggregator perspective which, in our case, is quite unique and directly reflects an actual aggregator in Denmark with incumbent market rules.

\changes We have extended our literature review in the manuscript (p.4, Section 1.2) with the following:

\nt{Similar studies exist in the literature albeit in different settings. For example, [20] uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in [21] although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In [22], it is shown how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves in the Nordics [23], in a microgrid [24], and in the Australian power grid [25].}

@@

\textbf{Comment 6:} \textit{Section 2 and 3: The math looks ok.}

\auth Thank you.

\textbf{Comment 7:} \textit{Section 4: The mFRR models also consider energy prices. Therefore, they also consider some load shifting. Please explain why the discussion is framed as load shifting vs mFRR since load shifting is apparently included in mFRR models. In my opinion, this part deserves a better explanation.}

\auth Thanks for this comment, which made us realized the original manuscript was not sufficiently clear. The energy price is only a variable in the load shifting optimization, not in the mFRR optimization. It was only included in the mFRR model eq.(2) for cost comparison to load shifting.

\changes To make it clear, we have changed several things: (i) Merged Figures 4 and 5 into one figure showing the variables for both mFRR and load shifting, respectively.(ii) revised section 2.2, 2.3, and 3.1 to accommodate those changes. Furthermore, eq.(2) has been changes such that the energy cost term is not present, but we mention it immediately after:

\begin{align}\label{eq:mFRRObjective}
  % & \underbrace{C(\rm{cost})}_{\rm{mFRR}} = - \underbrace{\sum_{h=1}^{24}
  \nt{\underbrace{\sum_{h=1}^{24}\lambda_{h}^{\rm{r}} p^{\rm{r}, \uparrow}_{h}}_{\textrm{Reservation payment}} + \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\uparrow}_{h}}_{\textrm{Activation payment}} - \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\downarrow}_{h}}_{\textrm{Rebound cost}} - \underbrace{ \sum_{h=1}^{24}  \lambda^{\rm{p}}s_{h}}_{\textrm{Penalty cost}}.
  }
\end{align}

\nt{
Note that an additional term containing the energy cost $\sum_{h=1}^{24} \lambda^{\rm{s}}_{h}P^{\rm{Base}}_{h}$ is removed from \eqref{eq:mFRRObjective}, since it is a fixed term in our setup. This term contains $\lambda^{\rm{s}}_{h}$ and $P^{\rm{Base}}_{h}$, which are both parameters, and therefore, can be omitted from the objective function. However, we will consider it when numerically comparing cost savings associated with two alternatives of mFRR provision and load shifting.
}

@@

[...]From a policy perspective, it is concerning if load shifting happens to be \nt{almost as} profitable \nt{as} mFRR provision.[...]

%%%%%%%%%%%%%%%%%
% REVIEWER 3
%%%%%%%%%%%%%%%%
\newpage
\section{Response to Reviewer 3}

\subsection{Summary Comment} \textit{I think the topic is quite interesting, the text is easy to read, and the English is quite good also. By reading the text I had some questions, that later I found as future works in the conclusions.}

\auth Thank you for your positive feedback and for highlighting the clarity and readability of our paper. We are glad to hear that you found the topic interesting and the language accessible. We also appreciate your engagement with our work and attention to details and inconsistencies in our references. Your main question regarding our scenario approach highlighted a need to clarify and justify this in more detail. We hope our response, outlining the details around our sensitivity analysis on the number of scenarios used, provides satisfactory.

@@

\subsection{Specific Comments}

Further, we address the Specific Comments in the following.

\textbf{Comment 1:} \textit{I think the paper is quite fine in general. My only concern is regarding the scenario generation (Section 3.2). Please, provide more details on the 5 days look back strategy.}

\auth Thanks for this pertinent comment. It is indeed a valid observation. First of all, the purpose of the ADMM strategy is to present a simple, i.e., fixed policy, that an aggregator can use to bid for entire 2022 without ever having to run a daily optimization. Such a policy is practical and appealing to real-life aggregators. On the contrary, the lookback policy might exhibit a better performance, especially in case of non-stationarity, but more complicated to implement and incorporate in daily operations. None of this was mentioned in our initial manuscript, hence we have added some sentences to clarify (please see below).

Second, during the analysis phase preceding this manuscript, a sensitivity analysis was made of both the ADMM strategy (based on scenarios from 2021) and for the lookback strategy (based on the last five days). The sensitivity analyses focused on the number of scenarios used in both cases and showed that ADMM performed the best when using at least 50 scenarios (with no improved performance using additional scenarios which we added as an appendix to the manuscript), and the same was the case for using five days in the lookback strategy. This has been clarified in the revised manuscript (please see below).

\nt{We use two different strategies for in-sample scenario generation: (\textit{i}) considering historical spot and balancing prices in DK2 in 2021.
%, with different cases where the number of scenarios $|\Omega|$ is 1, 5, 10, 20, 30, 40, 50, 100, and 250. 
(\textit{ii}) considering prices of the most recent five days (lookback strategy). This first scenario generation strategy represents the case where an aggregator relies on a simple, fixed policy without the need to run a daily optimization as opposed to the second  strategy, although it might exhibit a better performance in case the underlying uncertainty is non-stationary. The first  strategy uses 50 scenarios. Using a sensitivity analysis, we found out that increasing the number of scenarios beyond 50 does not improve the performance remarkably as explained in Appendix B. We made a similar observation for choosing five days in the second solution strategy.}

We then add Appendix B (including newly added Figure 8) as follows:

\nt{A sensitivity analysis was carried out to investigate the number of scenarios to use in the mFRR model using  2021 data. Fig. 8 shows the in-sample (IS) and out-of-sample (OOS) costs when increasing the number of scenarios used from 2021. It was eventually chosen to use 50 scenarios as this provided a satisfactory performance both IS and OOS --- a higher number of scenarios did not yield any particular improvement.}

### 2024-04-10 tex/paper_flex_v2.tex
Submission. Only missing final proof editing from EPSR editor.

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve:  Which One is More Appealing \textcolor{red}{to Thermostatically Controlled Loads in Denmark}?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  \textcolor{red}{distinct} forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to \textcolor{red}{Thermostatically Controlled Loads (TCLs)} in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the \textcolor{red}{TCL} in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative \textcolor{red}{TCL}, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be \textcolor{red}{almost as profitable as mFRR provision, although it could  be more} consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the \textcolor{red}{attraction of ancillary service provision  more obvious for TCLs in comparison to the load shifting alternative.}

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby \textcolor{red}{supermarket freezers as  Thermostatically Controlled Loads (TCLs)}  can shift consumption in time when spot (\textcolor{red}{also known as} day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this \textcolor{red}{purpose} as they constitute a big share of their energy consumption.

@@

\subsection{Research questions and literature review}
%
Using real data from Denmark, this work aims to investigate and answer the following three research questions: (\textit{i}) how can the flexibility of a supermarket freezer as a \textcolor{red}{TCL} be characterized?
(\textit{ii}) how much monetary benefit can this \textcolor{red}{TCL} earn by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the Danish mFRR market versus load shifting? Unlike the second and third questions, the first one has been addressed extensively in the literature. Therefore, the following literature review focuses on the first question, where we also differentiate our work from the current literature. To the best of our knowledge, there is no  work in the existing literature addressing the second and third questions.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. \textcolor{red}{These two} work\textcolor{red}{s} \textcolor{red}{provide a} benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a TCL is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\textcolor{red}{In \cite{schaperow2019simulation,chanpiwat2020using, moglen2020optimal}}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the U.S. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a trade-off is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

\textcolor{red}{Similar studies exist in the literature albeit in different settings. For example, \cite{iria2018trading} uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in \cite{la2021mixed} although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In \cite{nitsch2021economic}, it is shown how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves in the Nordics \cite{paridari2020flexibility}, in a microgrid \cite{mendieta2020primary}, and in the Australian power grid \cite{attarha2020network}}.

\subsection{Our contributions and outline}
%
To the best of our knowledge, no work has \textcolor{red}{thoroughly investigated and compared the monetary value and incentives for TCLs to} provide mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from \textcolor{red}{a supermarket's} perspective in Denmark. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a TCL. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper.  \textcolor{red}{Appendix A} presents the full model formulation. Finally, \textcolor{red}{Appendix B provides a sensitivity analysis on the number of scenarios.}

@@

There are several \textcolor{red}{distinct} ways to \textcolor{red}{exploit and} monetize flexibility from a \textcolor{red}{TCL}. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. 
For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

\subsection{Modeling TCL as a flexible resource}
%
TCLs are characterized by being controlled such that the temperature is kept at a specified setpoint. Examples include heat pumps, freezers, air condition units, etc. They constitute an important part of demand-side flexibility due to  inherent thermal inertia of such temperature-driven systems \textcolor{red}{\cite{koch2011modeling,hao2014aggregate}}.

We focus on freezers only, which are a very common type of TCLs. Specifically, we focus on a single freezer display in a Danish supermarket. Freezers are characterized by a large thermal inertia due to the frozen food, which makes them suitable for flexibility \textcolor{red}{provision}. On the other hand, there is a risk of food degradation when utilizing flexibility. Therefore, it is important to model the temperature dynamics in the freezer for a realistic and risk-aware estimation of its flexibility.

@@

\subsection{mFRR}\label{sec:mFRR}
%
Fig. \ref{fig:timeline_mfrr_variables} shows the timeline of the mFRR market in Denmark. There is a market for up-regulation only. One should note  that BRPs can choose \textit{not} to bid in the reserve market and  only bid in the real time for up- and down-regulation. We consider a case wherein the flexible demand delivers reservation through a BRP, since the payment is received for both reservation and activation. All prices considered are for the bidding zone DK2 (eastern Denmark).

@@

\begin{figure}[t]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{\textcolor{red}{Decision-making timeline and variables for the TCL when providing flexibility either in the form of mFRR services (green boxes) or load shifting (blue box).}}
    \label{fig:timeline_mfrr_variables}
\end{figure}

>>>>

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve:  Which One is More Appealing \textcolor{black}{to Thermostatically Controlled Loads in Denmark}?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  \textcolor{black}{distinct} forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to \textcolor{black}{Thermostatically Controlled Loads (TCLs)} in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the \textcolor{black}{TCL} in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative \textcolor{black}{TCL}, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be \textcolor{black}{almost as profitable as mFRR provision, although it could  be more} consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the \textcolor{black}{attraction of ancillary service provision  more obvious for TCLs in comparison to the load shifting alternative.}

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby \textcolor{black}{supermarket freezers as  Thermostatically Controlled Loads (TCLs)}  can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this \textcolor{black}{purpose} as they constitute a big share of their energy consumption.

@@

\subsection{Research questions and literature review}
%
Using real data from Denmark, this work aims to investigate and answer the following three research questions: (\textit{i}) how can the flexibility of a supermarket freezer as a \textcolor{black}{TCL} be characterized?
(\textit{ii}) how much monetary benefit can this \textcolor{black}{TCL} earn by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the Danish mFRR market versus load shifting? Unlike the second and third questions, the first one has been addressed extensively in the literature. Therefore, the following literature review focuses on the first question, where we also differentiate our work from the current literature. To the best of our knowledge, there is no  work in the existing literature addressing the second and third questions.

@@

In \cite{petersen2012eso2} and \cite{pedersen2013direct}, a complete white-box model of a supermarket refrigeration system is presented and validated against real-life data. It is also shown how such a system can provide demand response. \textcolor{black}{These two} work\textcolor{black}{s} \textcolor{black}{provide a} benchmark for any grey-box model of a supermarket refrigeration system. However, such an approach is hard to scale and requires complete knowledge of every refrigeration system. In \cite{pedersen2016improving}, a second-order model is used to model the food and air temperature in a freezer in an experimental setting. It is shown how food temperature has much slower dynamics than air temperature. In \cite{hao2014aggregate}, a simple first-order virtual battery model for a TCL is presented. This model also constitutes the starting point for modeling temperature dynamics in this work, as it has an intuitive interpretation. A similar bucket model is introduced in \cite{petersen2013taxonomy}.

@@

\textcolor{black}{In \cite{schaperow2019simulation,chanpiwat2020using, moglen2020optimal}}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the U.S. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a trade-off is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

\textcolor{black}{Similar studies exist in the literature albeit in different settings. For example, \cite{iria2018trading} uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in \cite{la2021mixed} although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In \cite{nitsch2021economic}, it is shown how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves in the Nordics \cite{paridari2020flexibility}, in a microgrid \cite{mendieta2020primary}, and in the Australian power grid \cite{attarha2020network}}.

\subsection{Our contributions and outline}
%
To the best of our knowledge, no work has \textcolor{black}{thoroughly investigated and compared the monetary value and incentives for TCLs to} provide mFRR services versus load shifting, although both have been studied extensively. This work aims to fill that gap by taking a holistic view on the incentives from \textcolor{black}{a supermarket's} perspective in Denmark. Furthermore, when stating the objective functions for mFRR, most studies have assumed a simplified market structure, and neglect the fact that the aggregator and the consumer are not necessarily the same entity. We discuss the consequences of this assumption in relation to the market structure in Denmark \cite{gade2022ecosystem}, and in the context of mFRR and load shifting.

@@

The rest of the paper is organized as follows. Section \ref{sec:monetizing_flex} describes modeling a supermarket freezer as a TCL. Thereafter, mFRR and load shifting are described in detail and their objective functions are stated. Section \ref{sec:OptimizationModel} presents the proposed stochastic optimization problem.
%Furthermore, the two solution strategies for solving the optimization problem for mFRR are introduced. 
Section \ref{sec:results} provides  results for a case study. Section \ref{sec:conclusion} concludes the paper.  \textcolor{black}{Appendix A} presents the full model formulation. Finally, \textcolor{black}{Appendix B provides a sensitivity analysis on the number of scenarios.}

@@

There are several \textcolor{black}{distinct} ways to \textcolor{black}{exploit and} monetize flexibility from a \textcolor{black}{TCL}. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. 
For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

\subsection{Modeling TCL as a flexible resource}
%
TCLs are characterized by being controlled such that the temperature is kept at a specified setpoint. Examples include heat pumps, freezers, air condition units, etc. They constitute an important part of demand-side flexibility due to  inherent thermal inertia of such temperature-driven systems \textcolor{black}{\cite{koch2011modeling,hao2014aggregate}}.

We focus on freezers only, which are a very common type of TCLs. Specifically, we focus on a single freezer display in a Danish supermarket. Freezers are characterized by a large thermal inertia due to the frozen food, which makes them suitable for flexibility \textcolor{black}{provision}. On the other hand, there is a risk of food degradation when utilizing flexibility. Therefore, it is important to model the temperature dynamics in the freezer for a realistic and risk-aware estimation of its flexibility.

@@

\subsection{mFRR}\label{sec:mFRR}
%
\textcolor{black}{The green boxes in Fig.\ref{fig:timeline_mfrr_variables} show the decision making timeline for the TCL to participate in the Danish mFRR market. The mFRR market in Denmark only trades up-regulation services.} One should note  that BRPs can choose \textit{not} to bid in the reserve market and  only bid in the real time for up- and down-regulation. We consider a case wherein the TCL delivers reservation through a BRP, since the payment is received for both reservation and activation. All prices considered are for the bidding zone DK2 (eastern Denmark).

@@

\begin{figure}[t]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{\textcolor{black}{Decision-making timeline and variables for the TCL when providing flexibility either in the form of mFRR services (green boxes) or load shifting (blue box).}}
    \label{fig:timeline_mfrr_variables}
\end{figure}
